---
title: "Getting Started with rarefun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rarefun}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The **rarefun** package provides a collection of functions for detecting and analyzing rare events in time series. 
This vignette demonstrates the main functionality of the package.

## Installation

You can install the development version of rarefun from GitHub:

```{r install}
# install.packages("devtools")
devtools::install_github("vlyubchich/rarefun")
```

```{r setup}
library(rarefun)
library(ggplot2)
```


## Anomaly Detection and Analysis

### Residual-based Anomaly Detection with BJsales Data

We'll use the built-in BJsales and BJsales.lead time series to demonstrate anomaly detection and comparison.

# Visualize time series and anomalies for BJsales
ggplot(res_sales$data, aes(x = time, y = value)) +
  geom_line(color = "#2c3e50", size = 0.7) +
  geom_point(data = subset(res_sales$data, is_anomaly_iforest),
       aes(x = time, y = value), color = "red", size = 2) +
  labs(title = "BJsales Time Series with Rare Events (Isolation Forest)",
    x = "Time", y = "Sales") +
  theme_minimal()

# Visualize time series and anomalies for BJsales.lead
ggplot(res_lead$data, aes(x = time, y = value)) +
  geom_line(color = "#2c3e50", size = 0.7) +
  geom_point(data = subset(res_lead$data, is_anomaly_iforest),
       aes(x = time, y = value), color = "red", size = 2) +
  labs(title = "BJsales.lead Time Series with Rare Events (Isolation Forest)",
    x = "Time", y = "Sales Lead") +
  theme_minimal()

  
```{r residuals-bjsales, eval=TRUE}
# Detect anomalies in BJsales
res_sales <- rare_residuals(BJsales)
# Detect anomalies in BJsales.lead
res_lead <- rare_residuals(BJsales.lead)

# Visualize anomalies in BJsales
ggplot(res_sales$data, aes(x = time, y = value)) +
  geom_line(color = "gray") +
  geom_point(aes(color = is_anomaly_iforest)) +
  labs(title = "Anomaly Detection in BJsales (Isolation Forest)",
    x = "Time", y = "Sales", color = "Anomaly") +
  scale_color_manual(values = c("gray", "red"), labels = c("Normal", "Anomaly")) +
  theme_minimal()

# Visualize anomalies in BJsales.lead
ggplot(res_lead$data, aes(x = time, y = value)) +
  geom_line(color = "gray") +
  geom_point(aes(color = is_anomaly_iforest)) +
  labs(title = "Anomaly Detection in BJsales.lead (Isolation Forest)",
    x = "Time", y = "Sales Lead", color = "Anomaly") +
  scale_color_manual(values = c("gray", "red"), labels = c("Normal", "Anomaly")) +
  theme_minimal()

# Compare anomaly labels using MCC
mcc_result <- mcc(res_sales$data$is_anomaly_iforest, res_lead$data$is_anomaly_iforest)
print(paste("MCC between BJsales and BJsales.lead anomalies:", round(mcc_result$mcc, 3)))
```


## Supporting Functions


### Isolation Forest Example (iris dataset)

The Isolation Forest algorithm is effective for detecting anomalies in high-dimensional data. We'll use the iris dataset and visualize results with ggplot2.

```{r iforest-iris, eval=TRUE}

# Use only numeric columns
iris_num <- iris[, 1:4]
result <- rare_iforest(iris_num, ntrees = 100, threshold = 0.6)
iris_plot <- iris_num
iris_plot$is_anomaly <- result$is_anomaly

ggplot(iris_plot, aes(x = Sepal.Length, y = Sepal.Width)) +
  labs(title = "Isolation Forest Anomaly Detection (iris)",
    x = "Sepal Length", y = "Sepal Width", color = "Anomaly") +
  scale_color_manual(values = c("blue", "red"), labels = c("Normal", "Anomaly")) +
  theme_minimal()
```


### DBSCAN Clustering Example (iris dataset)

DBSCAN (Density-Based Spatial Clustering of Applications with Noise) identifies anomalies as points that don't belong to any dense cluster. We'll use the iris dataset and visualize results with ggplot2.

```{r dbscan-iris, eval=TRUE}
data(iris)
iris_num <- iris[, 1:4]
result <- rare_dbscan(iris_num, minPts = 5)
iris_plot <- iris_num
iris_plot$is_anomaly <- result$is_anomaly
iris_plot$cluster <- factor(result$cluster)

ggplot(iris_plot, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = cluster, shape = is_anomaly)) +
  labs(title = "DBSCAN Anomaly Detection (iris)",
    x = "Sepal Length", y = "Sepal Width", color = "Cluster", shape = "Anomaly") +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = c(19, 17), labels = c("Normal", "Anomaly")) +
  theme_minimal()
```

### Kneedle Algorithm

The `kneedle()` function detects knee points in curves, useful for determining optimal thresholds:

```{r kneedle-example}
# Create curve with knee point
x <- 1:100
y <- 100 / (1 + exp(-(x - 50) / 10))

# Find knee point
knee <- kneedle(x, y)
print(paste("Knee point at x =", knee))

# Visualize
plot(x, y, type = "l", lwd = 2,
     main = "Knee Point Detection")
abline(v = knee, col = "red", lty = 2, lwd = 2)
```

### Spatial Point Matching

The `match_spatial_points()` function matches spatial points between datasets:

```{r spatial-example}
# Create two sets of spatial points
set.seed(789)
points1 <- data.frame(
  lon = runif(50, -80, -70),
  lat = runif(50, 35, 45)
)

points2 <- data.frame(
  lon = runif(30, -80, -70),
  lat = runif(30, 35, 45)
)

# Match points within distance threshold
matches <- match_spatial_points(points1, points2, max_dist = 0.5)
head(matches)
```


### Goodness-of-fit for Quantile Regression

The `gof_qr()` function tests goodness-of-fit for quantile regression models:

```{r gof-example}
# Fit quantile regression (requires quantreg package)
# library(quantreg)
# data(engel)
# fit <- rq(foodexp ~ income, data = engel, tau = 0.5)
# 
# # Test goodness-of-fit
# gof_result <- gof_qr(fit)
# print(gof_result)
```

### Partial Dependence Plot for Quantile Regression

The `partial_qrf()` function performs partial dependence analysis for quantile random forests:

```{r pqrf-example}
# Example with random forest (requires quantregForest package)
# library(quantregForest)
# set.seed(101)
# x <- matrix(rnorm(500), ncol = 5)
# y <- x[,1] + 0.5 * x[,2]^2 + rnorm(100)
# 
# # Fit model and compute partial dependence
# qrf_model <- quantregForest(x, y)
# partial <- partial_qrf(qrf_model, x, variable = 1)
```

## Best Practices

1. **Data Processing**: Always check for missing values and scale your data appropriately before applying anomaly detection methods.

2. **Parameter Tuning**: The effectiveness of anomaly detection methods depends on parameter choices. Consider cross-validation or domain knowledge to set parameters like `threshold`, `minPts`, or `ntrees`.

3. **Validation**: Always validate detected anomalies with domain expertise or additional data.

4. **Multiple Methods**: Consider using multiple detection methods and combining their results for more robust anomaly detection.

## Further Reading

For more details on specific functions, see their help pages:

```{r help}
?rare_iforest
?rare_dbscan
?rare_residuals
# etc.
```
