---
title: "Getting Started with rarefun"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        fig_caption: yes
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with rarefun}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    fig.width = 8,
    fig.height = 5,
    out.width = "90%",
    dpi = 150,
    message = FALSE
)
```

## Introduction

The package `rarefun` provides a collection of functions for detecting and analyzing rare events in time series. 
This vignette demonstrates the main functionality of the package.

You can install the development version of `rarefun` from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vlyubchich/rarefun")
```

```{r}
# Load the package
library(rarefun)

# Load packages for data manipulation
library(dplyr)

# Load packages for plotting
library(ggplot2)
theme_set(theme_minimal())
library(patchwork)
```


## Residual-based anomaly detection and analysis (non-seasonal time series)

### Minimal example

As a minimal example, we will use the built-in `BJsales` and `BJsales.lead` non-seasonal time series (Figure \@ref(fig:BJsales)).

```{r}
#| label: BJsales
#| fig-cap: "BJsales and BJsales.lead time series."

par(mfrow = c(1, 2))
plot(BJsales, 
     ylab = "BJsales", 
     xlab = "Time", 
     las = 1)
grid()
plot(BJsales.lead, 
     ylab = "BJsales.lead", 
     xlab = "Time", 
     las = 1)
grid()
par(mfrow = c(1, 1))
```

**First, detect anomalies** in each time series using the function `rare_residuals()`, which removes smooth trends and identifies rare events in the residuals.
Since the trends in the data are not very smooth, we will use a small span for the locally weighted regression (loess) smoother, for the smooth to be more local.
We use `ntrees = 100` for the isolation forest method to speed up the computation (increase `ntrees` for more consistent results).

```{r}
# Set seed for reproducibility
set.seed(123)

# Detect anomalies in BJsales
res_sales <- rare_residuals(BJsales, 
                            loess_args = list(span = 0.1),
                            iforest_args = list(ntrees = 100))

# Detect anomalies in BJsales.lead
res_lead <- rare_residuals(BJsales.lead, 
                           loess_args = list(span = 0.1),
                           iforest_args = list(ntrees = 100))
```

**Second, visualize the identified anomalies** (Figure \@ref(fig:rareBJsales)).

```{r}
#| label: rareBJsales
#| fig-cap: "Anomalies detected in BJsales and BJsales.lead using isolation forest."

# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with isolation forest",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with isolation forest",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1, axes = "collect"))
```

**Third, study the associations** between the rare events in these two time series.
Remember to use `ts = TRUE` so that the time series structure is taken into account when bootstrapping the Matthews correlation coefficient (MCC) p-values.
The `positive_class` argument defines the label for the rare events.
For example, the contemporaneous associations are as follows, resulting in non-significant MCC and $\chi^2$ tests:

```{r}
res0 <- mcc(res_sales$data$is_anomaly_iforest, 
            res_lead$data$is_anomaly_iforest,
            positive_class = "TRUE",
            ts = TRUE)
res0$mcc_bootstrap_pv
res0$chi_square_test$p.value
```

In fact, `BJsales.lead` is known to lead `BJsales` by about 3 time steps, so we explore associations between rare events at lag 3 (now displaying all outputs of the function):

```{r}
mcc(res_sales$data$is_anomaly_iforest, 
    dplyr::lag(res_lead$data$is_anomaly_iforest, 3),
    positive_class = "TRUE",
    ts = TRUE)
```

The results indicate a significant association between rare events in `BJsales.lead` at lag 3 and `BJsales`.

**Finally, extract the identified rare events** for `BJsales.lead` and optionally classify them as positive or negative anomalies:

```{r}
res_lead$data %>% 
    filter(is_anomaly_iforest) %>%
    mutate(type = ifelse(residual > 0, "positive", "negative")) %>%
    select(time, value, type)
```


### Extended analysis

Note that the function `rare_residuals()` also automatically applied DBSCAN method for detecting rare events (Figure \@ref(fig:rareBJsalesDBSCAN); the argument `method` controls the methods used for detecting rare events), for which a similar analysis using the function `mcc()` can be performed.

```{r}
#| label: rareBJsalesDBSCAN
#| fig-cap: "Anomalies detected in BJsales and BJsales.lead using DBSCAN."

# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with DBSCAN",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with DBSCAN",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1, axes = "collect"))
```

The `rare_residuals()` output provides anomaly scores for each method that can be used to customize the anomaly detection further. 
The scores range from 0 to 1, with higher values indicating more anomalous points.
Thus, we can adjust the threshold for the method to be more or less strict, without re-running the code.
For example, run a sensitivity analysis for several thresholds applied to the isolation forest outputs, counting the number of rare events for each threshold:

```{r}
thresholds <- seq(0.4, 0.9, by = 0.1)
results_list <- sapply(thresholds, function(thresh)
    sum(res_sales$data$score_iforest > thresh)
)
names(results_list) <- paste0("Threshold_", thresholds)
results_list
```

Also, the `rare_residuals()` output provides the original data (`value`) and residuals from the LOESS/STL smoother (`residual`), which can be further analyzed.

```{r}
#| label: trendBJsales
#| fig-cap: "BJsales trend and residuals."

# Visualize trend for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value - residual)) +
    geom_line() +
    labs(title = "BJsales trend",
         x = "Time", y = "BJsales")

# Visualize residuals for BJsales
p2 <- ggplot(res_sales$data, aes(x = time, y = residual)) +
    geom_line() +
    labs(title = "BJsales trend residuals",
         x = "Time", y = "BJsales")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1, axes = "collect"))
```

Note that if you would like to remove trends in a different way (for example, `diff(BJsales)`) or bypass the trend removal step, you can use the `rare_iforest()` and `rare_dbscan()` functions directly.
See the sections below for examples.


## Anomaly detection in seasonal time series

For seasonal time series, we can use the built-in `Annapolis` dataset, which contains 30 years (1995-2024) of daily maximum temperature (`tmax`) and precipitation (`pcp`) from Daymet for Annapolis, Maryland.

```{r}
# Load the Annapolis dataset
data(Annapolis)

# Quick summaries
head(Annapolis)
summary(Annapolis)
```

Visualize the temperature time series (Figure \@ref(fig:annapolistmax)).

```{r}
#| label: annapolistmax
#| fig-cap: "Daily maximum temperature in Annapolis, Maryland (1995-2024)."

ggplot(Annapolis, aes(x = date, y = tmax)) +
    geom_line(color = "gray50", linewidth = 0.3) +
    labs(x = "Date", y = "Maximum temperature (°C)")
```


### Residual-based anomalies

We can detect anomalies in seasonal time series by first removing the trend and seasonal components. 
The function `rare_residuals()` supports seasonal decomposition via STL (seasonal time series decomposition using loess), then applies the isolation forest and DBSCAN to detect anomalies in residuals.

```{r}
# Convert to ts object (frequency = 365 for daily data)
tmax_ts <- ts(Annapolis$tmax, 
              start = c(1995, 1, 1), 
              frequency = 365)

# Set seed for reproducibility
set.seed(456)

# Detect anomalies in monthly temperature
res_tmax <- rare_residuals(tmax_ts, 
                           seasonal = TRUE, 
                           period = frequency(tmax_ts),
                           stl_args = list(l.window = 365 * 2 + 1),
                           method = "iforest",
                           iforest_args = list(ntrees = 100, threshold = 0.6))
```

Visualize the detected anomalies (Figure \@ref(fig:raretmax)).
The plot shows the positive anomalies are often represented not by the highest temperatures but by unusually warm days in spring, while the negative anomalies are often the lowest temperatures observed in Annapolis.

```{r}
#| label: raretmax
#| fig-cap: "Anomalies detected in daily maximum temperature using isolation forest."

res_tmax$data %>%
  mutate(type = ifelse(residual > 0, "positive", "negative")) %>%
  ggplot(aes(x = time, y = value)) +
  geom_line(color = "gray50") +
  geom_point(data = . %>% filter(is_anomaly_iforest),
             aes(x = time, y = value, color = type), size = 2) +
  labs(x = "Date", y = "Temperature (°C)",
       color = "Anomaly type") +
  scale_color_manual(values = c("positive" = "tomato", "negative" = "cornflowerblue"))
```

We can also export anomalous temperatures to further examine them:

```{r}
# Extract anomalous months
anomalous_tmax <- res_tmax$data %>%
  # Replace time with the Date format (instead of decimal year)
  mutate(time = Annapolis$date) %>%
  filter(is_anomaly_iforest) %>%
  mutate(type = ifelse(residual > 0, "positive", "negative")) %>%
  select(time, value, residual, type)

head(anomalous_tmax, 10)
```

Visualize the trend-seasonal component and residuals (Figure \@ref(fig:stltmax)).
The residual plot should be centered around zero and show as little patterns as possible, indicating that the STL decomposition has effectively removed the trend and seasonal patterns.

```{r}
#| label: stltmax
#| fig-cap: "STL decomposition components for daily maximum temperature."

# Create a data frame with STL components
p1 <- ggplot(res_tmax$data, aes(x = time, y = value - residual)) +
    geom_line() +
    labs(title = "Trend + Seasonal components",
         x = "Year", y = "Temperature (°C)")

p2 <- ggplot(res_tmax$data, aes(x = time, y = residual)) +
    geom_line() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    labs(title = "Residuals after STL",
         x = "Date", y = "Residual (°C)")

print(p1 + p2 + plot_layout(ncol = 1, axes = "collect"))
```


### Seasonality shifts

Beyond isolated anomalies, we can look for shifts in seasonal patterns. 
For example, we might want to detect whether certain years have systematically warmer or cooler temperatures than the long-term climatology.


```{r}

```




Calculate annual anomalies relative to the long-term monthly climatology:

```{r}
# Calculate long-term monthly climatology
monthly_clim <- Annapolis %>%
    mutate(month = as.integer(format(date, "%m"))) %>%
    group_by(month) %>%
    summarize(tmax_clim = mean(tmax, na.rm = TRUE),
              .groups = "drop")

# Calculate anomalies for each day
Annapolis_anom <- Annapolis %>%
    mutate(month = as.integer(format(date, "%m")),
           year = as.integer(format(date, "%Y"))) %>%
    left_join(monthly_clim, by = "month") %>%
    mutate(tmax_anomaly = tmax - tmax_clim)

# Calculate annual mean anomalies
annual_anom <- Annapolis_anom %>%
    group_by(year) %>%
    summarize(mean_anomaly = mean(tmax_anomaly, na.rm = TRUE),
              .groups = "drop")
```

Visualize annual temperature anomalies (Figure \@ref(fig:annualanom)):

```{r}
#| label: annualanom
#| fig-cap: "Annual mean temperature anomalies in Annapolis, MD."

ggplot(annual_anom, aes(x = year, y = mean_anomaly)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 0.5) +
    labs(title = "Annual temperature anomalies with trend",
         x = "Year", y = "Temperature anomaly (°C)")
```

We can also examine seasonal shifts by comparing anomalies across different seasons:

```{r}
# Calculate seasonal anomalies
seasonal_anom <- Annapolis_anom %>%
    mutate(season = case_when(
        month %in% c(12, 1, 2) ~ "Winter",
        month %in% c(3, 4, 5) ~ "Spring",
        month %in% c(6, 7, 8) ~ "Summer",
        month %in% c(9, 10, 11) ~ "Fall"
    )) %>%
    group_by(year, season) %>%
    summarize(mean_anomaly = mean(tmax_anomaly, na.rm = TRUE),
              .groups = "drop")

# Visualize seasonal anomalies
ggplot(seasonal_anom, aes(x = year, y = mean_anomaly, color = season)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    facet_wrap(~season, ncol = 2) +
    labs(title = "Seasonal temperature anomalies",
         x = "Year", y = "Temperature anomaly (°C)",
         color = "Season") +
    theme(legend.position = "none")
```


### Heatwave-type events

The `rare_heatwaves()` function provides a wrapper around the `heatwaveR` package to detect discrete, prolonged high- or low-temperature events.

Detect heatwaves in the Annapolis temperature data:

```{r}
# Detect heatwaves using default settings
# (90th percentile threshold, 5-day minimum duration)
hw <- rare_heatwaves(Annapolis, 
                     date_col = "date", 
                     value_col = "tmax",
                     climatology_period = c("1995-01-01", "2024-12-30"),
                     pctile = 90,
                     min_duration = 5)

# Summary of detected events
cat("Number of heatwaves detected:", nrow(hw$events), "\n")
```

Examine the detected heatwaves:

```{r}
# Display information about detected heatwaves
if (nrow(hw$events) > 0) {
    hw$events %>%
        select(event_no, date_start, date_end, duration, 
               intensity_max, intensity_cumulative) %>%
        head(10)
}
```

Visualize the temperature time series with the 90th percentile threshold and detected heatwaves (Figure \@ref(fig:heatwaves)):

```{r}
#| label: heatwaves
#| fig-cap: "Detected heatwaves in Annapolis, MD (last 3 years shown). Black line: daily tmax, Blue line: climatology, Green dashed line: 90th percentile, Red filled areas: Heatwaves. Note the `heatwaveR` syntax for filled polygons."

# Plot the last 3 years for clarity
hw$data %>% 
    filter(date >= as.Date("2022-01-01")) %>% 
    ggplot(aes(x = date, y = tmax)) +
    heatwaveR::geom_flame(aes(y2 = thresh), n = 5, fill = "salmon") +
    geom_line(aes(y = seas), color = "blue", linewidth = 0.5) +
    geom_line(aes(y = thresh), color = "darkgreen", linewidth = 0.5, linetype = "dashed") +
    geom_line(color = "black", linewidth = 0.3) +
    labs(x = "Date", y = "Temperature (°C)")
```

We can also detect cold spells by setting `cold_spells = TRUE`:

```{r}
# Detect cold spells (below 10th percentile for 5+ days)
cs <- rare_heatwaves(Annapolis, 
                     date_col = "date", 
                     value_col = "tmax",
                     climatology_period = c("1995-01-01", "2024-12-30"),
                     pctile = 10,
                     min_duration = 5,
                     cold_spells = TRUE)

# Summary of detected cold-spells
cat("Number of cold spell events detected:", nrow(cs$events), "\n")

# Display the first few cold spells
if (nrow(cs$events) > 0) {
    cs$events %>%
        select(event_no, date_start, date_end, duration, 
               intensity_max, intensity_cumulative) %>%
        head(10)
}
```


```{r}
#| label: coldspells
#| fig-cap: "Detected cold spells in Annapolis, MD (last 3 years shown). Black line: daily tmax, Blue line: climatology, Green dashed line: 10th percentile, Blue filled areas: cold spells. Note the default `ggplot2` syntax for filled polygons."

# Plot the last 3 years for clarity
cs$data %>% 
    filter(date >= as.Date("2022-01-01")) %>% 
    ggplot(aes(x = date, y = tmax)) +
    geom_line(aes(y = seas), color = "blue", linewidth = 0.5) +
    geom_line(aes(y = thresh), color = "darkgreen", linewidth = 0.5, linetype = "dashed") +
    geom_line(color = "black", linewidth = 0.3) +
    geom_ribbon(data = cs$data %>% filter(date >= as.Date("2022-01-01"), event),
        aes(x = date, ymin = tmax, ymax = thresh), fill = "cornflowerblue", color = NA) +
    labs(x = "Date", y = "Temperature (°C)")
```


## Supporting functions

### Isolation forest

The isolation forest algorithm can be applied separately using the `rare_iforest()` function.
For example, apply it to the `BJsales` time series detrended by taking consecutive differences:

```{r}
x <- as.matrix(diff(BJsales), ncol = 1)
res_diffsales <- rare_iforest(x, ntrees = 100)
```

Count the number of detected anomalies (`TRUE`) at the default threshold of 0.5:

```{r}
table(res_diffsales$is_anomaly)
```

Report times of the detected anomalies:

```{r}
time(diff(BJsales))[res_diffsales$is_anomaly] 
```

We can also apply the algorithm to **multivariate data.**
For example, apply it to the differenced `BJsales` and `BJsales.lead` time series:

```{r}
x <- cbind(diff(BJsales), diff(BJsales.lead))
res_diffsales2 <- rare_iforest(x, ntrees = 100)
table(res_diffsales2$is_anomaly)
```

See the help page `?rare_iforest` for other examples.


### DBSCAN clustering 

The DBSCAN clustering algorithm is adapted to label the noise cluster (observations not belonging to any cluster) as anomalies.
It can be applied separately using the function `rare_dbscan()`, similarly to the isolation forest example above.
For example, apply it to the `BJsales` time series detrended by taking consecutive differences:

```{r}
x <- as.matrix(diff(BJsales), ncol = 1)
res_diffsales <- rare_dbscan(x)
```

Count the number of detected anomalies (`TRUE`):

```{r}
table(res_diffsales$is_anomaly)
```

Report times of the detected anomalies:

```{r}
time(diff(BJsales))[res_diffsales$is_anomaly] 
```

We can also apply the algorithm to **multivariate data** (each variable is re-scaled automatically with the default setting `scale = TRUE`).
For example, apply it to the differenced `BJsales` and `BJsales.lead` time series:

```{r}
x <- cbind(diff(BJsales), diff(BJsales.lead))
res_diffsales2 <- rare_dbscan(x)
table(res_diffsales2$is_anomaly)
```

The DBSCAN output also includes the assigned clusters, which can be plotted as in Figure \@ref(fig:dbscan).

```{r}
#| label: dbscan
#| fig-cap: "DBSCAN clustering results for differenced BJsales and BJsales.lead."

x %>% 
  as.data.frame() %>% 
  mutate(cluster = res_diffsales2$cluster) %>%
  ggplot(aes(x = `diff(BJsales)`, 
             y = `diff(BJsales.lead)`, 
             color = as.factor(cluster))) +
  geom_point(size = 2) +
  labs(color = "Cluster")
```

See the help page `?rare_dbscan` for other examples.


### Kneedle algorithm

The function `kneedle()` detects knee points in curves, useful for determining optimal thresholds such as in elbow methods. 
A common application is identifying the optimal number of principal components in PCA by finding the "elbow" in the scree plot.

```{r}
# Perform PCA on mtcars dataset
pca_result <- prcomp(mtcars, scale. = TRUE)

# Extract variance explained by each component
variance_explained <- pca_result$sdev^2
pc_number <- seq_along(variance_explained)

# Find the knee point coordinates
(knee <- kneedle(pc_number, variance_explained))

# The x-value is the optimal number of components
n_components <- knee["x"]
```

```{r}
#| label: kneedle
#| fig-cap: "PCA scree plot with knee point detection."

# Visualize scree plot with knee point
plot(pc_number, variance_explained, type = "b", lwd = 2,
     xlab = "Principal component",
     ylab = "Variance explained",
     las = 1)
grid()
points(knee["x"], knee["y"], col = "red", pch = 19, cex = 1.5)
legend("topright", 
       legend = paste("Knee at PC", n_components),
       col = "red", pch = 19)
```

In the package `rarefun`, the `kneedle()` function is used internally in the function `rare_dbscan()` to help determine optimal value for the parameter `eps`.


### Spatial point matching

The function `match_spatial_points()` matches spatial points between datasets:

```{r spatial-example}
# Create two sets of spatial points, with sample sizes n1 and n2
n1 <- 50
n2 <- 30

set.seed(123)
pointsA <- data.frame(
    lon = runif(n1, -80, -70),
    lat = runif(n1, 35, 40),
    id = paste0("A_", 1:n1)
)
pointsB <- data.frame(
    lon = runif(n2, -80, -70),
    lat = runif(n2, 35, 40),
    id = paste0("B_", 1:n2)
)

# Match points within the distance threshold 50 km
matches <- match_spatial_points(pointsA, pointsB, max_dist = 50000)
head(matches)
```

The function tries to match points from `pointsB` to `pointsA`, hence the output size matches the number of elements in the dataset `pointsA`:

```{r}
nrow(matches) == nrow(pointsA)
```


### Goodness-of-fit for quantile regression

The function `gof_qr()` calculates goodness-of-fit metrics for quantile regression models.

(add example?)


### Partial dependence plots for quantile regression

The function `partial_qrf()` plots partial dependence plots for quantile random forests.

(add example?)


## Best practices

1. **Data Processing**: Always check for missing values and scale your data appropriately before applying anomaly detection methods.

2. **Parameter Tuning**: The effectiveness of anomaly detection methods depends on parameter choices. Consider cross-validation, sensitivity analysis, or domain knowledge to set parameters like `threshold`, `minPts`, or `ntrees`.

3. **Validation**: Always validate detected anomalies with domain expertise or additional data.

4. **Multiple Methods**: Consider using multiple detection methods and combining their results for more robust anomaly detection.


## Further reading

For more details on specific functions, see their help pages:

```{r, eval = FALSE}
?rare_iforest
?rare_dbscan
?rare_residuals
# etc.
```
