---
title: "Getting Started with rarefun"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        fig_caption: yes
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with rarefun}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    fig.width = 8,
    fig.height = 5,
    out.width = "90%",
    dpi = 150,
    message = FALSE
)
```

## Introduction

The package `rarefun` provides a collection of functions for detecting and analyzing rare events in time series. 
This vignette demonstrates the main functionality of the package.

You can install the development version of `rarefun` from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vlyubchich/rarefun")
```

```{r}
# Load the package
library(rarefun)

# Load packages for data manipulation
library(dplyr)

# Load packages for plotting
library(ggplot2)
theme_set(theme_minimal())
library(patchwork)
```


## Residual-based anomaly detection and analysis (non-seasonal time series)

### Minimal example

As a minimal example, we will use the built-in `BJsales` and `BJsales.lead` non-seasonal time series (Figure \@ref(fig:BJsales)).

```{r}
#| label: BJsales
#| fig-cap: "BJsales and BJsales.lead time series."

par(mfrow = c(1, 2))
plot(BJsales, 
     ylab = "BJsales", 
     xlab = "Time", 
     las = 1)
grid()
plot(BJsales.lead, 
     ylab = "BJsales.lead", 
     xlab = "Time", 
     las = 1)
grid()
par(mfrow = c(1, 1))
```

**First, detect anomalies** in each time series using the function `rare_residuals()`, which removes smooth trends and identifies rare events in the residuals.
Since the trends in the data are not very smooth, we will use a small span for the LOESS smoother, for the smooth to be more local.
We use `ntrees = 100` for the isolation forest method to speed up the computation (increase `ntrees` for more consistent results).

```{r}
# Set seed for reproducibility
set.seed(123)

# Detect anomalies in BJsales
res_sales <- rare_residuals(BJsales, 
                            loess_args = list(span = 0.1),
                            iforest_args = list(ntrees = 100))

# Detect anomalies in BJsales.lead
res_lead <- rare_residuals(BJsales.lead, 
                           loess_args = list(span = 0.1),
                           iforest_args = list(ntrees = 100))
```

**Second, visualize the identified anomalies** (Figure \@ref(fig:rareBJsales)).

```{r}
#| label: rareBJsales
#| fig-cap: "Anomalies detected in BJsales and BJsales.lead using isolation forest."

# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with isolation forest",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with isolation forest",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

**Third, study the associations** between the rare events in these two time series.
Remember to use `ts = TRUE` so that the time series structure is taken into account when bootstrapping the Matthews correlation coefficient (MCC) p-values.
The `positive_class` argument defines the label for the rare events.
For example, the contemporaneous associations are as follows, resulting in non-significant MCC and $\chi^2$ tests:

```{r}
res0 <- mcc(res_sales$data$is_anomaly_iforest, 
            res_lead$data$is_anomaly_iforest,
            positive_class = "TRUE",
            ts = TRUE)
res0$mcc_bootstrap_pv
res0$chi_square_test$p.value
```

In fact, `BJsales.lead` is known to lead `BJsales` by about 3 time steps, so we explore associations between rare events at lag 3 (now displaying all outputs of the function):

```{r}
mcc(res_sales$data$is_anomaly_iforest, 
    dplyr::lag(res_lead$data$is_anomaly_iforest, 3),
    positive_class = "TRUE",
    ts = TRUE)
```

The results indicate a significant association between rare events in `BJsales.lead` at lag 3 and `BJsales`.

**Finally, extract the identified rare events** for `BJsales.lead` and optionally classify them as positive or negative anomalies:

```{r}
res_lead$data %>% 
    filter(is_anomaly_iforest) %>%
    mutate(type = ifelse(residual > 0, "positive", "negative")) %>%
    select(time, value, type)
```


### Extended analysis

Note that the function `rare_residuals()` also automatically applied DBSCAN method for detecting rare events (Figure \@ref(fig:rareBJsalesDBSCAN); the argument `method` controls the methods used for detecting rare events), for which a similar analysis using the function `mcc()` can be performed.

```{r}
#| label: rareBJsalesDBSCAN
#| fig-cap: "Anomalies detected in BJsales and BJsales.lead using DBSCAN."

# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with DBSCAN",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with DBSCAN",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

The `rare_residuals()` output provides anomaly scores for each method that can be used to customize the anomaly detection further. 
The scores range from 0 to 1, with higher values indicating more anomalous points.
Thus, we can adjust the threshold for the method to be more or less strict, without re-running the code.
For example, run a sensitivity analysis for several thresholds applied to the isolation forest outputs, counting the number of rare events for each threshold:

```{r}
thresholds <- seq(0.4, 0.9, by = 0.1)
results_list <- sapply(thresholds, function(thresh)
    sum(res_sales$data$score_iforest > thresh)
)
names(results_list) <- paste0("Threshold_", thresholds)
results_list
```

Also, the `rare_residuals()` output provides the original data (`value`) and residuals from the LOESS/STL smoother (`residual`), which can be further analyzed.

```{r}
#| label: trendBJsales
#| fig-cap: "BJsales trend and residuals."

# Visualize trend for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value - residual)) +
    geom_line() +
    labs(title = "BJsales trend",
         x = "Time", y = "BJsales")

# Visualize residuals for BJsales
p2 <- ggplot(res_sales$data, aes(x = time, y = residual)) +
    geom_line() +
    labs(title = "BJsales trend residuals",
         x = "Time", y = "BJsales")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

Note that if you would like to remove trends in a different way (for example, `diff(BJsales)`) or bypass the trend removal step, you can use the `rare_iforest()` and `rare_dbscan()` functions directly.
See the sections below for examples.


## Anomaly detection in seasonal time series

(Example with seasonal temp data)

### Residual-based anomalies

### Seasonality shifts

### Heatwave-type events






## Supporting functions

### Isolation forest

The isolation forest algorithm can be applied separately using the `rare_iforest()` function.
For example, apply it to the `BJsales` time series detrended by taking consecutive differences:

```{r}
x <- as.matrix(diff(BJsales), ncol = 1)
res_diffsales <- rare_iforest(x, ntrees = 100)
```

Count the number of detected anomalies (`TRUE`) at the default threshold of 0.5:

```{r}
table(res_diffsales$is_anomaly)
```

Report times of the detected anomalies:

```{r}
time(diff(BJsales))[res_diffsales$is_anomaly] 
```

We can also apply the algorithm to **multivariate data.**
For example, apply it to the differenced `BJsales` and `BJsales.lead` time series:

```{r}
x <- cbind(diff(BJsales), diff(BJsales.lead))
res_diffsales2 <- rare_iforest(x, ntrees = 100)
table(res_diffsales2$is_anomaly)
```

See the help page `?rare_iforest` for other examples.


### DBSCAN clustering 

The DBSCAN clustering algorithm is adapted to label the noise cluster (observations not belonging to any cluster) as anomalies.
It can be applied separately using the function `rare_dbscan()`, similarly to the isolation forest example above.
For example, apply it to the `BJsales` time series detrended by taking consecutive differences:

```{r}
x <- as.matrix(diff(BJsales), ncol = 1)
res_diffsales <- rare_dbscan(x)
```

Count the number of detected anomalies (`TRUE`):

```{r}
table(res_diffsales$is_anomaly)
```

Report times of the detected anomalies:

```{r}
time(diff(BJsales))[res_diffsales$is_anomaly] 
```

We can also apply the algorithm to **multivariate data** (each variable is re-scaled automatically with the default setting `scale = TRUE`).
For example, apply it to the differenced `BJsales` and `BJsales.lead` time series:

```{r}
x <- cbind(diff(BJsales), diff(BJsales.lead))
res_diffsales2 <- rare_dbscan(x)
table(res_diffsales2$is_anomaly)
```

The DBSCAN output also includes the assigned clusters, which can be plotted as in Figure \@ref(fig:dbscan).

```{r}
#| label: dbscan
#| fig-cap: "DBSCAN clustering results for differenced BJsales and BJsales.lead."

x %>% 
  as.data.frame() %>% 
  mutate(cluster = res_diffsales2$cluster) %>%
  ggplot(aes(x = `diff(BJsales)`, 
             y = `diff(BJsales.lead)`, 
             color = as.factor(cluster))) +
  geom_point(size = 2) +
  labs(color = "Cluster")
```

See the help page `?rare_dbscan` for other examples.


### Kneedle algorithm

The function `kneedle()` detects knee points in curves, useful for determining optimal thresholds such as in elbow methods. 
A common application is identifying the optimal number of principal components in PCA by finding the "elbow" in the scree plot.

```{r}
# Perform PCA on mtcars dataset
pca_result <- prcomp(mtcars, scale. = TRUE)

# Extract variance explained by each component
variance_explained <- pca_result$sdev^2
pc_number <- seq_along(variance_explained)

# Find the knee point coordinates
(knee <- kneedle(pc_number, variance_explained))

# The x-value is the optimal number of components
n_components <- knee["x"]
```

```{r}
#| label: kneedle
#| fig-cap: "PCA scree plot with knee point detection."

# Visualize scree plot with knee point
plot(pc_number, variance_explained, type = "b", lwd = 2,
     xlab = "Principal component",
     ylab = "Variance explained",
     las = 1)
grid()
points(knee["x"], knee["y"], col = "red", pch = 19, cex = 1.5)
legend("topright", 
       legend = paste("Knee at PC", n_components),
       col = "red", pch = 19)
```

In the package `rarefun`, the `kneedle()` function is used internally in the function `rare_dbscan()` to help determine optimal value for the parameter `eps`.


### Spatial point matching

The function `match_spatial_points()` matches spatial points between datasets:

```{r spatial-example}
# Create two sets of spatial points, with sample sizes n1 and n2
n1 <- 50
n2 <- 30

set.seed(123)
pointsA <- data.frame(
    lon = runif(n1, -80, -70),
    lat = runif(n1, 35, 40),
    id = paste0("A_", 1:n1)
)
pointsB <- data.frame(
    lon = runif(n2, -80, -70),
    lat = runif(n2, 35, 40),
    id = paste0("B_", 1:n2)
)

# Match points within the distance threshold 50 km
matches <- match_spatial_points(pointsA, pointsB, max_dist = 50000)
head(matches)
```

The function tries to match points from `pointsB` to `pointsA`, hence the output size matches the number of elements in the dataset `pointsA`:

```{r}
nrow(matches) == nrow(pointsA)
```


### Goodness-of-fit for quantile regression

The function `gof_qr()` calculates goodness-of-fit metrics for quantile regression models.

(add example?)


### Partial dependence plots for quantile regression

The function `partial_qrf()` plots partial dependence plots for quantile random forests.

(add example?)


## Best practices

1. **Data Processing**: Always check for missing values and scale your data appropriately before applying anomaly detection methods.

2. **Parameter Tuning**: The effectiveness of anomaly detection methods depends on parameter choices. Consider cross-validation, sensitivity analysis, or domain knowledge to set parameters like `threshold`, `minPts`, or `ntrees`.

3. **Validation**: Always validate detected anomalies with domain expertise or additional data.

4. **Multiple Methods**: Consider using multiple detection methods and combining their results for more robust anomaly detection.


## Further reading

For more details on specific functions, see their help pages:

```{r, eval = FALSE}
?rare_iforest
?rare_dbscan
?rare_residuals
# etc.
```
