---
title: "Getting Started with rarefun"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        fig_caption: yes
vignette: >
%\VignetteIndexEntry{Getting Started with rarefun}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
    editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE
)
```

## Introduction

The package `rarefun` provides a collection of functions for detecting and analyzing rare events in time series. 
This vignette demonstrates the main functionality of the package.

## Installation

You can install the development version of `rarefun` from GitHub:

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vlyubchich/rarefun")
```

```{r setup}
# Load the package
library(rarefun)

# Load packages for data manipulation
library(dplyr)

# Load packages for plotting
library(ggplot2)
theme_set(theme_minimal())
library(patchwork)
```


## Residual-based anomaly detection and analysis (non-seasonal time series)

### Minimal example

As a minimal example, we will use the built-in `BJsales` and `BJsales.lead` non-seasonal time series (\@ref(fig:BJsales)).

```{r BJsales, fig.cap="BJsales and BJsales.lead time series."}
par(mfrow = c(1, 2))
plot(BJsales, 
     ylab = "BJsales", 
     xlab = "Time", 
     las = 1)
grid()
plot(BJsales.lead, 
     ylab = "BJsales.lead", 
     xlab = "Time", 
     las = 1)
grid()
```

First, detect anomalies in each time series using the function `rare_residuals()`, which removes smooth trends and identifies rare events in the residuals.
Since the trends in the data are not very smooth, we will use a small span for the LOESS smoother, for the smooth to be more local.
We use `ntrees = 100` for the isolation forest method to speed up the computation (increase `ntrees` for more consistent results).

```{r}
# Set seed for reproducibility
set.seed(123)

# Detect anomalies in BJsales
res_sales <- rare_residuals(BJsales, 
                            loess_args = list(span = 0.1),
                            iforest_args = list(ntrees = 100))

# Detect anomalies in BJsales.lead
res_lead <- rare_residuals(BJsales.lead, 
                           loess_args = list(span = 0.1),
                           iforest_args = list(ntrees = 100))
```

Second, visualize the identified anomalies (\@ref(fig:rareBJsales)).

```{r rareBJsales, fig-cap="Anomalies detected in BJsales and BJsales.lead using isolation forest."}
# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with isolation forest",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_iforest),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with isolation forest",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

Third, we can study the associations between the rare events in these two time series.
Remember to use `ts = TRUE` so that the time series structure is taken into account when bootstrapping the Matthews correlation coefficient (MCC) p-values.
For example, the contemporaneous associations are as follows, resulting in non-significant MCC and $\chi^2$ tests:

```{r}
res0 <- mcc(res_sales$data$is_anomaly_iforest, 
            res_lead$data$is_anomaly_iforest,
            ts = TRUE)
res0$mcc_bootstrap_pv
res0$chi_square_test$p.value
```

In fact, `BJsales.lead` is known to lead `BJsales` by about 3 time steps, so we explore associations between rare events at lag 3 (now displaying all outputs of the function):

```{r}
mcc(res_sales$data$is_anomaly_iforest, 
    dplyr::lag(res_lead$data$is_anomaly_iforest, 3),
    ts = TRUE)
```

The results indicate a significant association between rare events in `BJsales.lead` at lag 3 and `BJsales`.
The code below extracts the identified rare events for `BJsales.lead` and classifies them as positive or negative anomalies:

```{r}
res_lead$data %>% 
    filter(is_anomaly_iforest) %>%
    mutate(type = ifelse(residual > 0, "positive", "negative")) %>%
    select(time, value, type)
```


### Extended analysis

Note that the function `rare_residuals()` also automatically applied DBSCAN method for detecting rare events (\@ref(fig:rareBJsalesDBSCAN); the argument `method` controls the methods used for detecting rare events), for which a similar analysis using the function `mcc()` can be performed.

```{r rareBJsalesDBSCAN, fig-cap="Anomalies detected in BJsales and BJsales.lead using DBSCAN."}
# Visualize time series and anomalies for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_sales$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales with rare events identified with DBSCAN",
         x = "Time", y = "BJsales")

# Visualize time series and anomalies for BJsales.lead
p2 <- ggplot(res_lead$data, aes(x = time, y = value)) +
    geom_line(color = "gray50") +
    geom_point(data = subset(res_lead$data, is_anomaly_dbscan),
               aes(x = time, y = value), color = "red", size = 2) +
    labs(title = "BJsales.lead with rare events identified with DBSCAN",
         x = "Time", y = "BJsales.lead")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

The `rare_residuals()` output provides anomaly scores for each method that can be used to customize the anomaly detection further. 
The scores range from 0 to 1, with higher values indicating more anomalous points.
Thus, we can adjust the threshold for the method to be more or less strict, without re-running the code.
For example, run a sensitivity analysis for several thresholds applied to the isolation forest outputs, counting the number of rare events for each threshold:

```{r}
thresholds <- seq(0.4, 0.9, by = 0.1)
results_list <- sapply(thresholds, function(thresh)
    sum(res_sales$data$score_iforest > thresh)
)
names(results_list) <- paste0("Threshold_", thresholds)
results_list
```

Also, the `rare_residuals()` output provides the original data (`value`) and residuals from the LOESS/STL smoother (`residual`), which can be further analyzed.

```{r trendBJsales, fig-cap="Anomalies detected in BJsales and BJsales.lead using DBSCAN."}
# Visualize trend for BJsales
p1 <- ggplot(res_sales$data, aes(x = time, y = value - residual)) +
    geom_line() +
    labs(title = "BJsales trend",
         x = "Time", y = "BJsales")

# Visualize residuals for BJsales
p2 <- ggplot(res_sales$data, aes(x = time, y = residual)) +
    geom_line() +
    labs(title = "BJsales trend residuals",
         x = "Time", y = "BJsales")

# Combine plots
print(p1 + p2 + plot_layout(ncol = 1))
```

Note that if you would like to remove trends in a different way (for example, `diff(BJsales)`) or bypass the trend removal step, you can use the `rare_iforest()` and `rare_dbscan()` functions directly.
See the sections below for examples.


## Residual-based anomaly detection (seasonal time series)




## Supporting Functions


### Isolation forest example

The Isolation Forest algorithm is effective for detecting anomalies in high-dimensional data. We'll use the iris dataset and visualize results with ggplot2.

```{r iforest-iris}

# Use only numeric columns
iris_num <- iris[, 1:4]
result <- rare_iforest(iris_num, ntrees = 100, threshold = 0.6)
iris_plot <- iris_num
iris_plot$is_anomaly <- result$is_anomaly

ggplot(iris_plot, aes(x = Sepal.Length, y = Sepal.Width)) +
    labs(title = "Isolation Forest Anomaly Detection (iris)",
         x = "Sepal Length", y = "Sepal Width", color = "Anomaly") +
    scale_color_manual(values = c("blue", "red"), labels = c("Normal", "Anomaly")) +
    theme_minimal()
```


### DBSCAN Clustering Example (iris dataset)

DBSCAN (Density-Based Spatial Clustering of Applications with Noise) identifies anomalies as points that don't belong to any dense cluster. We'll use the iris dataset and visualize results with ggplot2.

```{r dbscan-iris}
data(iris)
iris_num <- iris[, 1:4]
result <- rare_dbscan(iris_num, minPts = 5)
iris_plot <- iris_num
iris_plot$is_anomaly <- result$is_anomaly
iris_plot$cluster <- factor(result$cluster)

ggplot(iris_plot, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point(aes(color = cluster, shape = is_anomaly)) +
    labs(title = "DBSCAN Anomaly Detection (iris)",
         x = "Sepal Length", y = "Sepal Width", color = "Cluster", shape = "Anomaly") +
    scale_color_brewer(palette = "Set1") +
    scale_shape_manual(values = c(19, 17), labels = c("Normal", "Anomaly")) +
    theme_minimal()
```

### Kneedle Algorithm

The `kneedle()` function detects knee points in curves, useful for determining optimal thresholds:

```{r kneedle-example}
# Create curve with knee point
x <- 1:100
y <- 100 / (1 + exp(-(x - 50) / 10))

# Find knee point
knee <- kneedle(x, y)
print(paste("Knee point at x =", knee))

# Visualize
plot(x, y, type = "l", lwd = 2,
     main = "Knee Point Detection")
abline(v = knee, col = "red", lty = 2, lwd = 2)
```

### Spatial Point Matching

The `match_spatial_points()` function matches spatial points between datasets:

```{r spatial-example}
# Create two sets of spatial points
set.seed(789)
points1 <- data.frame(
    lon = runif(50, -80, -70),
    lat = runif(50, 35, 45)
)

points2 <- data.frame(
    lon = runif(30, -80, -70),
    lat = runif(30, 35, 45)
)

# Match points within distance threshold
matches <- match_spatial_points(points1, points2, max_dist = 0.5)
head(matches)
```


### Goodness-of-fit for Quantile Regression

The `gof_qr()` function tests goodness-of-fit for quantile regression models:

```{r gof-example}
# Fit quantile regression (requires quantreg package)
# library(quantreg)
# data(engel)
# fit <- rq(foodexp ~ income, data = engel, tau = 0.5)
# 
# # Test goodness-of-fit
# gof_result <- gof_qr(fit)
# print(gof_result)
```

### Partial Dependence Plot for Quantile Regression

The `partial_qrf()` function performs partial dependence analysis for quantile random forests:

```{r pqrf-example}
# Example with random forest (requires quantregForest package)
# library(quantregForest)
# set.seed(101)
# x <- matrix(rnorm(500), ncol = 5)
# y <- x[,1] + 0.5 * x[,2]^2 + rnorm(100)
# 
# # Fit model and compute partial dependence
# qrf_model <- quantregForest(x, y)
# partial <- partial_qrf(qrf_model, x, variable = 1)
```

## Best Practices

1. **Data Processing**: Always check for missing values and scale your data appropriately before applying anomaly detection methods.

2. **Parameter Tuning**: The effectiveness of anomaly detection methods depends on parameter choices. Consider cross-validation or domain knowledge to set parameters like `threshold`, `minPts`, or `ntrees`.

3. **Validation**: Always validate detected anomalies with domain expertise or additional data.

4. **Multiple Methods**: Consider using multiple detection methods and combining their results for more robust anomaly detection.

## Further Reading

For more details on specific functions, see their help pages:

```{r help}
?rare_iforest
?rare_dbscan
?rare_residuals
# etc.
```
